Management induced changes of soil organic carbon on global croplands

### 2022-07-20 SOCbudgetPaper/SOCbudget.Rmd
updated manuscript

  Soil organic carbon (SOC), one of the largest terrestrial carbon (C) stocks on Earth, has been depleted by anthropogenic land-cover change and agricultural management. However, the latter has so far not been well represented in global C stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatially explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with a reduced-complexity model based on the IPCC Tier 2 method to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around $34.6\,\unit{GtC}$ relative to a counterfactual hypothetical natural state in 1975. Within the period 1975--2010 this SOC debt continued to expand by $5\,\unit{GtC}$ ($0.14\,\unit{GtC\,yr^{-1}}$) to around 39.6 GtC. However, accounting for historical management led to $2.1\,\unit{GtC}$ less ($0.06\,\unit{GtC\,yr^{-1}}$) emissions than under the assumption of constant management. We also find that management decisions have influenced the historical SOC trajectory most strongly by residue returning, indicating that increasing SOC sequestration by biomass retention may be a promising negative emissions technique. The reduced-complexity SOC model may allow to simulate management-induced SOC sequestration also within computationally demanding integrated (land-use) assessment modeling. 

@@

    ```{=latex}
  \begin{figure}[H]
    \includegraphics[width=16cm]{../ResultNotebooks/Output/Images/reg_comparisonfigure} 
    \caption{Modelled as well as data based estimation for climate zone specific SOC stock in GtC for the first 30 cm of soil aggregated over all land area: SoilGrids, GSOC and WISE do not consider changes over time and rely on soil profile data gather over a long period of time, which makes it hard to pinpoint a specific year to these SOC estimations. In this context they will be compared to modelled data (LPJmL4, this study) for the year 2010. PNV denotes the potential natural vegetation state without considering human cropping activities, calculated as reference stock within our model. We use the climate zone specification of the IPCC \citep{eggleston_ipcc_2006}.}
    \label{fig:SOCreg}
  \end{figure}
  ```

>>>>

  Soil organic carbon (SOC), one of the largest terrestrial carbon (C) stocks on Earth, has been depleted by anthropogenic land-cover change and agricultural management. However, the latter has so far not been well represented in global C stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatially explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with a reduced-complexity model based on the IPCC Tier 2 method to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around $34.6\,\unit{GtC}$ relative to a counterfactual hypothetical natural state in 1975. Within the period 1975--2010 this SOC debt continued to expand by $5\,\unit{GtC}$ ($0.14\,\unit{GtC\,yr^{-1}}$) to around 39.6 GtC. However, accounting for historical management led to $2.1\,\unit{GtC}$ less ($0.06\,\unit{GtC\,yr^{-1}}$) emissions than under the assumption of constant management. We also find that management decisions have influenced the historical SOC trajectory most strongly by residue returning, indicating that SOC enhancement by biomass retention may be a promising negative emissions technique. The reduced-complexity SOC model may allow to simulate management-induced SOC enhancement also within computationally demanding integrated (land-use) assessment modeling. 

@@

    ```{=latex}
  \begin{figure}[H]
    \includegraphics[width=12cm]{../ResultNotebooks/Output/Images/scenario_init.png} 
    \caption{Global $\Delta SOC$ for different SOC initialization choices in the start year 1901: Starting in 1901 with steady-state SOC under vegetation for all land-use types without any human cropping activities (Initial-natveg) lead to a smaller $\Delta SOC$ in 1975 and a steeper increase till 2010, as compared to initialzing with steady-state SOC stocks under historic land-use (Initial-spinup1510). On the other hand, assuming all SOC to be in land-use specific steady-state already in 1901 (Initial-eq) leads to the opposite effect of an already higher $\Delta SOC$ in 1975 and a less steeper increase till 2010. However, the difference between the dynamic historical management assumption (histManagement) compared to constant management assumption from 1975 on (constManagement) are of a similar size for all initialization choices.}
    \label{fig:SOCinit}
  \end{figure}
  ```

    ```{=latex}
  \begin{figure}[H]
    \includegraphics[width=16cm]{../ResultNotebooks/Output/Images/reg_comparisonfigure.png} 
    \caption{Modelled as well as data based estimation for climate zone specific SOC stock in GtC for the first 30 cm of soil aggregated over all land area: SoilGrids, GSOC and WISE do not consider changes over time and rely on soil profile data gather over a long period of time, which makes it hard to pinpoint a specific year to these SOC estimations. In this context they will be compared to modelled data (LPJmL4, this study) for the year 2010. PNV denotes the potential natural vegetation state without considering human cropping activities, calculated as reference stock within our model. We use the climate zone specification of the IPCC \citep{eggleston_ipcc_2006}.}
    \label{fig:SOCreg}
  \end{figure}
  ```

### 2022-07-20 SOCbudgetPaper/discussion.Rmd
updated manuscript

Our spatially explicit time series data set of agricultural management is based on country-specific FAO production and cropland statistics [@faostat_faostat_2016] as well as 0.5 degree land-use data from LUH2 [@hurtt_harmonization_2020]. Starting from these two sources, we derive a harmonized and consistent data set for the major C flows within the agricultural system (\@ref(fig:FlowFig)) using a mass balance approach from the IPCC guidelines Vol. 4 [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019] and other auxiliary data sets [e.g. @porwollik_generating_2019].

@@

@@

Generally, limiting the analysis to the first 30 cm of the soil profile follows the IPCC guidelines [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019] and assumes that most of the SOC dynamics happen in the topsoil. In this regard several aspects are strongly simplified within our approach. Firstly, distribution of carbon inputs into different soil layers are neglected and all carbon inputs are allocated to the topsoil. This particularly overestimates SOC stocks in the first 30 cm of soil below deeper rooting vegetation, which is certainly the case for most of the woody natural vegetated areas. Second, changes to the subsoil due to tillage are neglected. As Powlson et al. [-@powlson_limited_2014] have shown, the subsoil can be make a large difference in evaluating total SOC losses or gains for no-tillage systems. No-tillage effects may seem larger than they actually are if only topsoil is considered. SOC transfers to deeper soil layers under tillage might enhance subsoil SOC compared to no-till practices. Finally, organic soils (like peat- and wetlands) and drained cropland areas are not explicitly considered and emissions from these cropland areas are thus likely substantially underestimated.

@@

The analysis of SOC stock gains and losses is complex and has several dimensions as climatic and anthropogenic effects overlap. There is broad consensus that land conversion to cropland has caused substantial C emissions over the historical period [e.g. @frielingstein_glob_carbon_2020]. There is uncertainty with respect to the overall size of these emissions from different methods and reference points and with respect to the contribution of cropland and agricultural management to these emissions. In order to mitigate greenhouse gas emissions, it is essential to stop the decline of SOC stocks or even transform cropland management to sequester atmospheric C in cropland soils [@minasny_soil_2017]. Defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on cropland as the difference between a potential natural state and the state under human interventions [see @pugh_simulated_2015], we find that global cropland has acted as a emissions source since 1975. Annual C loss rates of 0.2 per 1000 C still have the opposite trend as the promoted 4 per 1000 C sequestration rate target [@minasny_soil_2017]. Dedicated efforts to increase cropland SOC are thus necessary, as management improvements at historical rates are not enough to counteract ongoing SOC degradation on cropland. Yet our study also shows the substantial impact of changing management on the development of SOC debt (Fig. \@ref(fig:SOCscen)).

@@

Furthermore, Sanderman et al. [-@sanderman_soil_2017] modeled historical trends based on agricultural land expansion without considering SOC variations due to time-variant agricultural management. Pugh et al. [-@pugh_simulated_2015] considered management effects like tillage and incorporation of residues in stylized and static scenarios only, so that they could not account for historical management effects on SOC dynamics. Their study moreover concludes that yield gains (by 18% in their simulations) do not lead to a substantial decline in SOC debt (less than 1% change). Historical yield increases, however, are often estimated to be well above 50% [@pellegrini_crop_2018; @ray_recent_2012; @rudel_agricultural_2009]. While we find substantially larger effects of productivity gains than the 1% reported by Pugh et al. [-@pugh_simulated_2015], this is not sufficient to compensate SOC losses from moderate global cropland expansion of around 11% between 1974 and 2010.

@@

There is also ample potential for further improved SOC management. As shown in Fig. \@ref(fig:FlowFig), approximately one fifth of total annual C sequestration by crops is lost through soils (0.8 GtC per year). However, even larger losses occur at the end of the food supply chain (1.2 GtC year), at the soil surface (1.4 GtC), during residue burning (0.3 Gt C) and with manure management (0.2). Improved management could include, firstly, a circular flow from the food supply chain back to soils. Waste composting or excreta recycling could represent a major additional C input to cropland soils [@brenzinger_organic_2018]. Secondly, soil carbon sequestration techniques [@smith_soil_2016], deep ploughing [@alcantara_deep_2016] or the transformation of C inputs to more recalcitrant biochar [@woolf_sustainable_2010] may transfer larger parts of the biomass at the litter soil barrier into permanent soil pools. Thirdly, reducing the share of residue burning and improved manure recycling could further increase C inputs. Finally, other carbon-accumulating practices, such as the cultivation of cover crops [@poeplau_carbon_2015; @porwollik_covercrops_2022] and agroforestry [@lorenz_soil_2014] could increase total C sequestration on cropland.

>>>>

Our spatially explicit time series data set of agricultural management is based on country-specific FAO production and cropland statistics [@faostat_faostat_2016] as well as 0.5 degree land-use data from LUH2 [@hurtt_harmonization_2020]. Starting from these two sources, we derive a harmonized and consistent data set for the major C flows within the cropping system (\@ref(fig:FlowFig)) using a mass balance approach from the IPCC guidelines Vol. 4 [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019] and other auxiliary data sets [e.g. @porwollik_generating_2019].

@@

The initialization of SOC stocks, however, is important for the size of the SOC debt and its change over time, since the presence and size of legacy fluxes effect these values strongly (see Fig. \@ref(fig:SOCinit)). According to our sensitivity analysis the SOC debt depending on the initialization choice has been between $33.3\,\unit{GtC}$ and $50.7\,\unit{GtC}$, with our best guess at $39.6\,\unit{GtC}$. Concurrently, our results indicate, that the impact of the dynamic agricultural management is robust to the initialization of SOC stocks.

@@

Generally, limiting the analysis to the first 30 cm of the soil profile follows the IPCC guidelines [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019] and assumes that most of the SOC dynamics happen in the topsoil. In this regard several aspects are strongly simplified within our approach. Firstly, distribution of carbon inputs into different soil layers are neglected and all carbon inputs are allocated to the topsoil. This particularly overestimates SOC stocks in the first 30 cm of soil below deeper rooting vegetation, which is certainly the case for most of the woody natural vegetated areas. Second, changes to the subsoil are neglected, which is most important for tillage effects. Other management practices might be more equally effecting top- and subsoil as they do not directly the relocates carbon vertically. As Powlson et al. [-@powlson_limited_2014] have shown, the subsoil can be make a large difference in evaluating total SOC losses or gains for no-tillage systems. No-tillage effects may seem larger than they actually are if only topsoil is considered. SOC transfers to deeper soil layers under tillage might enhance subsoil SOC compared to no-till practices. However, the effect of no-till to the subsoil as well as its overall importance as a SCS measure is still debated [@ogle_climate_2019]. Finally, organic soils (like peat- and wetlands) and drained cropland areas are not explicitly considered and emissions from these cropland areas are thus likely substantially underestimated.

@@

The analysis of SOC stock gains and losses is complex and has several dimensions as climatic and anthropogenic effects overlap. There is broad consensus that land conversion to cropland has caused substantial C emissions over the historical period [e.g. @frielingstein_glob_carbon_2020]. There is uncertainty with respect to the overall size of these emissions from different methods and reference points and with respect to the contribution of cropland and agricultural management to these emissions. In order to mitigate greenhouse gas emissions, it is essential to stop the decline of SOC stocks or even transform cropland management to sequester atmospheric C in cropland soils [@minasny_soil_2017]. Defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on cropland as the difference between a potential natural state and the state under human interventions [see @pugh_simulated_2015], we find that global cropland has acted as a emissions source since 1975. Annual C loss rates of 0.2 per 1000 C still have the opposite trend as the promoted 4 per 1000 C sequestration rate target [@minasny_soil_2017]. Dedicated efforts to increase cropland SOC are thus necessary, as management improvements at historical rates are not enough to counteract ongoing SOC degradation on cropland. Yet our study also shows the substantial impact of changing management on the development of SOC debt (Fig. \@ref(fig:SOCdebt) and Fig. \@ref(fig:SOCscen)).

@@

Furthermore, Sanderman et al. [-@sanderman_soil_2017] modeled historical trends based on agricultural land expansion without considering SOC variations due to time-variant agricultural management. Pugh et al. [-@pugh_simulated_2015] considered management effects like tillage and incorporation of residues in stylized and static scenarios only, so that they could not account for historical management effects on SOC dynamics. Their study moreover concludes that yield gains (by 18% in their simulations) do not lead to a substantial decline in SOC debt (less than 1% SOC increase). Historical yield increases, however, are estimated to be well above 50% [@pellegrini_crop_2018; @ray_recent_2012; @rudel_agricultural_2009] and often lead to an increase in below- and above-ground residue biomass inputs to the soil. While we find substantially larger SOC increase in response to productivity gains than the 1% reported by Pugh et al. [-@pugh_simulated_2015], this is not sufficient to compensate SOC losses from global cropland expansion of around 11% between 1974 and 2010.

@@

There is also ample potential for further improved SOC management. As shown in Fig. \@ref(fig:FlowFig), the annual SOC respiration (1.3 GtC per year) is slightly above one quarter of total annual net C uptake by crops (4.6 GtC per year). C compounds have to be respired by soil organisms to maintain basic soil functions and regulate the nutrient cycle, which often leaves limited options to decrease C losses via SOC respiration [@janzen_soil_2006].
However, similar C losses occur at the end of the food supply chain (1.2 GtC per year), at the soil surface (1.5 GtC per year), and smaller but still considerable during residue burning (0.2 GtC per year) and within animal waste management systems (0.2 GtC per year). Improved management could include, firstly, a circular flow from the food supply chain back to soils. Waste composting or excreta recycling could represent a major additional C input to cropland soils [@brenzinger_organic_2018]. Secondly, soil carbon sequestration techniques [@smith_soil_2016], deep ploughing [@alcantara_deep_2016] or the transformation of C inputs to more recalcitrant biochar [@woolf_sustainable_2010] may transfer larger parts of the biomass at the litter soil barrier into permanent soil pools. Thirdly, reducing the share of residue burning and improved manure recycling could further increase C inputs. Finally, other carbon-accumulating practices, such as the cultivation of cover crops [@poeplau_carbon_2015; @porwollik_covercrops_2022] and agroforestry [@lorenz_soil_2014] could increase total C sequestration on cropland.
